\clearpage
\begin{centering}
\textbf{ACKNOWLEDGEMENTS}\\
\vspace{\baselineskip}
\end{centering}

%Insert your dedication text here

\noindent I would like to thank my parents, Mike and Trish, for their love and support throughout my life.
Thank you so much for instilling the curiousity and dedication in me which made this possible.
My sisters, brothers, nieces, nephews, and cousins deserve my appreciation as well.

\hfill

\noindent I would like to sincerely thank my supervisor, Prof. Sherrill, for his guidance and support.
His belief in me provided encouragement throughout this entire process. I would also like to
thank Dr. Chow for co-advising me. The insights I learned from his courses made much of this work
possible. Also, I would like to thank Dr. Smith, who, without his mentorship and guidance, none
of this would have been possible. Collaborating with him was truly inspiring.

\hfill

\noindent I greatly appreciate the support of groupmates within the Sherrill Lab. Brandon, thank you for
your superb automation skills and willingness to run tests for me. Lori, thank you for letting me bug you often about
how to compile Psi4. Thank you Dom, Assim, Constance, Carlos, and Yi for the support and fun times. 

\hfill

\noindent Last but not least, I would like to thank my girlfriend, Andrea, whose emotional and intellectual support
kept me above water throughout graduate school.

\hfill

\noindent This research was supported by NSF grant ACI-1609842. 
\clearpage
%\pagenumbering{gobble}  %remove page number on summary page
